# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
Read the csv file into a dataframe and provide a summary of the data.
```{r}
df <- read.csv(file="./activity.csv",
               colClasses=c("integer","character","integer"))
cat("Number of rows and columns:",dim(df))
cat("Data frame overview")
str(df)
```

Keep only complete observations
```{r}
df <- na.omit(df)
cat("Number of complete rows and columns:",dim(df))
```


## What is mean total number of steps taken per day?
Calculate the total number steps per day.
```{r}
steps.day <- by(df$steps,df$date,sum)
head(steps.day)
```

```{r HistTotalDailySteps,fig.width=6,fig.height=5}
hist(steps.day,
     main="Histogram of Total Number of Steps",
     xlab="Total Number of Daily Steps")
````

Calculate mean and median total number of steps per day.
```{r}
mean.total.steps <- mean(steps.day)
median.total.steps <- median(steps.day)
```

* Mean total number of steps taken per day is **`r mean.total.steps`**
* Median total number of steps taken per day is **`r median.total.steps`**


## What is the average daily activity pattern?
Now calculate the average interval activty over the course of a day.
```{r}
## calculate average number of steps in an interval over all days
interval.steps <- by(df$steps,df$interval,mean)
head(interval.steps)
```

Plot of activity over daily interval
```{r ActivityPattern,fig.width=6,fig.height=5}
## convert character interval names to numeric values
interval <- as.integer(rownames(interval.steps))

## generate line chart
plot(interval,interval.steps,type="l",
     main="Average Daily Activity Pattern",
     ylab="Average number of steps",
     xlab="Interval")
```

Determine interval with the maximum value.
```{r}
max.steps.idx <-which(interval.steps == max(interval.steps))
```

Five-minute interval "`r names(max.steps.idx)`" has the maximum average number of steps of `r interval.steps[max.steps.idx]`.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
