# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
Download data from the web and unzip raw data.  Read the csv file into a dataframe and provide a summary of the data.
```{r}
###
# Data retrieved from
#  https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
#
# The data is downloaded and unzipped into the current working direcory to create
# activity file.
###

## Download the file
file.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" 
my.file <- "./repdata_data_activity.zip" 

setInternet2(TRUE) 
download.file(file.url,my.file) 
unzip(my.file)

raw.df <- read.csv(file="./activity.csv",
               colClasses=c("integer","character","integer"))
cat("Data frame overview\n")
cat("Data retrieved on",format(Sys.time(),"%m/%d/%Y %H:%M:%S"))
str(raw.df)
```

Keep only complete observations
```{r}
complete.df <- na.omit(raw.df)
cat("Number of complete rows:",dim(complete.df)[1])
```


## What is mean total number of steps taken per day?
Calculate the total number steps per day.
```{r}
steps.day <- by(complete.df$steps,complete.df$date,sum)

```

```{r HistTotalDailySteps,fig.width=6,fig.height=5}
hist(steps.day,
     main="Histogram of Total Number of Steps",
     xlab="Total Number of Daily Steps")
````

Calculate mean and median total number of steps per day.
```{r}
mean.total.steps <- mean(steps.day)
median.total.steps <- median(steps.day)
```

* Mean total number of steps taken per day is **`r sprintf("%.3f",mean.total.steps)`**
* Median total number of steps taken per day is **`r median.total.steps`**


## What is the average daily activity pattern?
Now calculate the average interval activty over the course of a day.
```{r}
## calculate average number of steps in an interval over all days
interval.steps <- tapply(complete.df$steps,complete.df$interval,mean)
head(interval.steps)
```

Plot of activity over daily interval
```{r ActivityPattern,fig.width=6,fig.height=5}
## convert character interval names to numeric values
interval <- as.integer(rownames(interval.steps))

## generate line chart
plot(interval,interval.steps,type="l",
     main="Average Daily Activity Pattern",
     ylab="Average number of steps",
     xlab="Interval")
```

Determine interval with the maximum value.
```{r}
max.steps.idx <- which(interval.steps == max(interval.steps))
```

Five-minute interval "`r names(max.steps.idx)`" has the maximum average number of steps of `r interval.steps[max.steps.idx]`.

## Imputing missing values
There are `r sum(!complete.cases(raw.df))` observations with missing values.

The strategy for imputing missing values is to use the median steps for each 
interval across the days.  First we calculate the median for each interval.

```{r}
interval.median.steps <- tapply(raw.df$steps,
                                     raw.df$interval,
                                     median,na.rm=TRUE)
```
Here is a sample of the interval medians that will be used for imputation.
```{r}
## sample of median values
n <- as.integer(0.75*length(interval.median.steps))
interval.median.steps[n:(n+5)]

```

The procedure for imputing missing values is to find missing values for each 
interval and to replace any missing values with the median calculated for that 
interval.

```{r}
## get valid values for all intervals
intervals <- unique(raw.df$interval)

## make copy of raw data
imputed.df <- raw.df

## cycle thru the intervals and replace missing values
for (intvl in intervals) {
    ## get median value for specific interval
    median.value <- interval.median.steps[names(interval.median.steps) == intvl]
    
    ## replace missing missing values
    imputed.df[imputed.df$interval == intvl & is.na(imputed.df$steps), "steps"] <- median.value
}
```

Summary of the original raw data.
```{r echo=FALSE}
summary(raw.df)
```
Summary of the data after imputing median values for missing values.
```{r echo=FALSE}
summary(imputed.df)
```

This is the histogram with imputed values.
```{r}
steps2.day <- by(imputed.df$steps,imputed.df$date,sum)
```
```{r HistTotalDailyStepsImputed,fig.width=6,fig.height=5}
hist(steps2.day,
     main="Histogram of Total Number of Steps (Imputed)",
     xlab="Total Number of Daily Steps")
```

We re-calculate mean and median total number of steps per day after imputing
missing values.
```{r}
mean.total.steps2 <- mean(steps2.day)
median.total.steps2 <- median(steps2.day)
```
Revised values after imputation are
* Mean total number of steps taken per day is **`r sprintf("%.3f",mean.total.steps2)`**
* Median total number of steps taken per day is **`r median.total.steps2`**

The effect of imputing mising values resulted in lowering both the mean and median
values.

## Are there differences in activity patterns between weekdays and weekends?

We now determine if there is a difference between activity during the week versus 
the weekend.

First we enrich the data with an indicator for **weekday** (Monday through Friday) 
or **weekend** (Saturday and Sunday).
```{r ,warning=FALSE}
imputed.df$weekday.ind <- 
    factor(ifelse(weekdays(as.Date(imputed.df$date)) %in% c("Saturday","Sunday"),
                                 "weekend","weekday"))

## overview of enriched data
str(imputed.df)

## calculate average number of steps in an interval over all days
interval2.steps <- data.frame(tapply(imputed.df$steps,
                      list(imputed.df$interval,imputed.df$weekday.ind),mean))

interval2.steps$interval <- as.integer(rownames(interval2.steps))


## reshape the data from plotting
library(reshape2)
interval2.df <- melt(interval2.steps,id=c("interval"),
                     measure.vars=c("weekend","weekday"))

names(interval2.df)[2:3] <- c("timeperiod","steps")
```
Overview of the enriched data for plotting.
```{r}
str(interval2.df)
head(interval2.df,3)
tail(interval2.df,3)
```


The following plot compares the activity for weekdays vs weekends.
```{r WeekdayActivityComparision,fig.width=6,fig.height=5,warning=FALSE}
library(lattice)
xyplot(steps ~ interval | timeperiod, data=interval2.df,
       type="l",
       ylab="Number of steps",
       layout=c(1,2))
````

